---
title: "Bayesian Adaptive Phase II Oncology Trial"
subtitle: "Operating characteristics, real-data validation, and a mock SAP"
author: "Cris Taylor"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
    fig-width: 8
    fig-height: 5
  pdf:
    toc: true
    toc-depth: 3
    geometry: margin=1in
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
suppressPackageStartupMessages({
  library(here)
  library(dplyr)
  library(tibble)
  library(knitr)
  library(flextable)
})
source(here("R", "00_setup.R"))
oc       <- readRDS(file.path(PATH_SIMS, "oc_table.rds"))
design   <- readRDS(file.path(PATH_SIMS, "design_params.rds"))
cox_obj  <- readRDS(file.path(PATH_SIMS, "tcga_cox_model.rds"))
km_tbl   <- read.csv(file.path(PATH_TBL, "tcga_km_summary.csv"))
cox_tbl  <- read.csv(file.path(PATH_TBL, "tcga_cox_results.csv"))
aft_tbl  <- read.csv(file.path(PATH_TBL, "tcga_bayes_aft_summary.csv"))
```

## Executive summary {.unnumbered}

We simulated 12,000 Phase II oncology trials (1,000 per scenario × 2 designs)
under 6 prior scenarios spanning hazard ratios (HR) from harmful (1.15) to
strongly beneficial (0.55). A Bayesian response-adaptive design with one
event-driven interim futility look (30% information time under the design
alternative HR = 0.70, fires when 12 events accumulate) was compared
against a fixed group-sequential O'Brien-Fleming design that applies the
identical final-stage z-boundary. The adaptive design controlled Type I
error at **0.019** (vs 0.021 for fixed; both below the 0.025 one-sided
target), stopped early for futility in **48% of trials under harmful HR**
and **36% under the null** (5–8% expected sample-size savings in those
scenarios), and ceded **0.6–5.6 percentage points of power** to the fixed
design across non-null effects. A parallel survival analysis on
n = 1,002 TCGA-BRCA patients exercised the same survival toolkit on real
data: Cox PH and a Bayesian Weibull AFT directionally agree that
hormone-receptor positive status is protective (HR ≈ 0.58–0.70) and that
each decade of age increases hazard (HR ≈ 1.17–1.28), with a Schoenfeld
test flagging a proportional-hazards violation that motivates the
parametric AFT cross-check.

## 1. Introduction

Phase II oncology programs face a power–efficiency trade-off: fixed designs
with adequate power are routinely overpowered for harmful or null effects,
exposing more patients than necessary to an ineffective experimental
therapy. Bayesian adaptive designs with response-adaptive randomization
(RAR) and futility interim looks are now recognized in the FDA
*Adaptive Designs for Clinical Trials of Drugs and Biologics* (2019)
guidance as appropriate tools, provided that operating characteristics are
demonstrated by simulation under all plausible scenarios.

This report quantifies the operating characteristics of one such design
against a fixed alternative for a hypothetical time-to-progression endpoint,
and demonstrates the same survival analytic toolkit on real breast-cancer
data from TCGA-BRCA.

## 2. Estimand (ICH E9(R1))

The primary estimand follows the ICH E9(R1) five-attribute structure.
A "treatment policy" strategy is used for the most operationally common
intercurrent event (treatment discontinuation), which is consistent with
the FDA's most-frequent recommendation for exploratory Phase II oncology
trials.

```{r estimand-table}
estimand <- tibble::tribble(
  ~Attribute,                ~Specification,
  "Treatment",               "Experimental therapy vs standard-of-care control, both administered until disease progression, unacceptable toxicity, or withdrawal of consent.",
  "Population",              "Patients with the target indication meeting protocol-defined eligibility.",
  "Endpoint (variable)",     "Time from randomization to progression or death from any cause (PFS).",
  "Intercurrent events",     "Treatment discontinuation handled via treatment-policy strategy: events occurring after discontinuation are included in the analysis.",
  "Population-level summary","Hazard ratio (experimental vs control) at the end of follow-up."
)
flextable::flextable(estimand) |>
  flextable::theme_booktabs() |>
  flextable::autofit() |>
  flextable::fontsize(size = 10, part = "all")
```

## 3. Trial designs compared

| Feature | Fixed design | Adaptive design |
|---|---|---|
| Max sample size | 120 (60 / arm) | 120 |
| Allocation | 1:1 throughout | 1:1 until interim; RAR after |
| Interim look | None | Event-driven: fires when 12 observed events accumulate (30% information under H1) |
| Interim futility rule | — | Stop if P(HR < 0.7 \| data) < 0.20 |
| Final test statistic | Cox PH `z = -log_hr / se` | Cox PH `z = -log_hr / se` |
| Decision boundary | OBF z = 1.969 (final stage) | OBF z = 1.969 (final stage; no efficacy stopping at interim) |
| RAR scheme | — | Thompson-style: `alloc_treat = max(0.2, min(0.8, sqrt(P(treat better))))`, refit every 20 enrollees post-interim |

The adaptive design's interim Bayesian model is an exponential survival
model with weakly informative priors centered on the baseline truth:
`λ_c ~ Gamma(2, 80)` (E[λ_c] = 0.025/month, matching the data-generating
control hazard) and `log HR ~ N(0, 1)`.

## 4. Simulation methods

**Data-generating model.** Patients enroll at rate 12/month
(Exp(1/12) inter-arrival times). Event times are exponential with control
monthly hazard 0.025 (annual hazard 0.30) and treatment hazard scaled by
the scenario's true HR. Administrative censoring is applied at 24 months
from study start.

**Scenarios.** Six prior scenarios spanning the relevant HR range:

```{r scenarios-table}
oc |>
  dplyr::filter(design == "fixed") |>
  dplyr::transmute(scenario, hr_true,
                   description = c("Treatment harms (HR > 1)",
                                   "No effect (null, alpha control)",
                                   "Modest treatment benefit",
                                   "Moderate treatment benefit",
                                   "Strong treatment benefit",
                                   "Very strong treatment benefit")) |>
  flextable::flextable() |>
  flextable::theme_booktabs() |>
  flextable::autofit()
```

**Reproducibility.** All randomness is seeded from `CONFIG$simulation$seed`
(20260513) with per-sim seeds derived as
`seed + sim_id * 10 + as.integer(factor(design))`. Each (scenario × design)
cell runs 1,000 sims. `furrr::future_pmap()` parallelizes over four worker
processes; the full 12,000-sim run completes in ~100 s wall time.

## 5. Operating characteristics

```{r oc-table}
oc |>
  dplyr::mutate(
    `Pr(reject)` = sprintf("%.3f", reject_rate),
    `95% CI`     = sprintf("(%.3f, %.3f)", reject_ci_lo, reject_ci_hi),
    `Pr(futility)` = ifelse(is.na(p_futility_stop), "—",
                             sprintf("%.2f", p_futility_stop)),
    `E[N]`       = sprintf("%.1f", e_n),
    `E[events]`  = sprintf("%.1f", e_events),
    `Mean HR`    = sprintf("%.3f", mean_hr_est),
    `Bias log-HR` = sprintf("%+.3f", bias_log_hr)
  ) |>
  dplyr::select(scenario, design, hr_true, `Pr(reject)`, `95% CI`,
                `Pr(futility)`, `E[N]`, `E[events]`, `Mean HR`, `Bias log-HR`) |>
  flextable::flextable() |>
  flextable::theme_booktabs() |>
  flextable::autofit() |>
  flextable::fontsize(size = 9, part = "all")
```

### 5.1 Power curve

::: {layout-ncol=1}
![Rejection rate by scenario and design with exact 95% binomial CIs.
The adaptive design (red) is 1–5 percentage points below fixed (blue) at
every non-null effect, the cost of stopping for futility under the
alternative.](../outputs/figures/oc_power_curve.pdf)
:::

### 5.2 Expected sample size

::: {layout-ncol=1}
![The adaptive design enrolls ~10 fewer patients on average under harmful
and null scenarios (futility stops cut enrollment short); savings shrink
as the true effect strengthens.](../outputs/figures/oc_expected_n.pdf)
:::

### 5.3 Probability of futility stop

::: {layout-ncol=1}
![Adaptive design only. Futility stop probability is highest under
harmful (17%) and null (12%) and drops to 3% under very-strong effect —
the design correctly distinguishes signal from noise even on the limited
information available at 50% enrollment.](../outputs/figures/oc_futility_stop.pdf)
:::

### 5.4 Summary heatmap

::: {layout-ncol=1}
![All six metrics tiled by scenario and faceted by design. Side-by-side
comparison makes the operational vs statistical trade-off visible at a
glance.](../outputs/figures/oc_summary_heatmap.pdf)
:::

### 5.5 Group-sequential boundary cross-validation

The R `{rpact}` design (k = 2 stages, O'Brien-Fleming alpha spending,
one-sided α = 0.025, 80% power) yields the following stage-wise
boundaries:

```{r design-boundaries}
design$boundaries_tbl |>
  dplyr::mutate(
    across(c(efficacy_z_boundary, futility_z_boundary,
             cumulative_alpha_spent, cumulative_beta_spent),
           ~ sprintf("%.4f", .))
  ) |>
  flextable::flextable() |>
  flextable::theme_booktabs() |>
  flextable::autofit()
```

The companion `sas/seqdesign.sas` produces the same boundaries via PROC
SEQDESIGN with identical alpha/beta-spending settings, verifying the
design specification across the R and SAS implementations.

![Alpha-spending function (and futility-spending boundary) from rpact.](../outputs/figures/alpha_spending.pdf)

## 6. Real-data case study: TCGA-BRCA

The same survival analytic pipeline (KM + Cox PH + Bayesian AFT) is
exercised on overall survival in n = 1,002 TCGA-BRCA patients (97 events,
median follow-up 487 days), stratified by hormone-receptor (HR) status.
This demonstrates that the toolkit works on real, messier data; it does
**not** validate the simulator's data-generating model, since the
endpoint (overall survival in breast cancer) and population differ
substantially from the simulator's hypothetical time-to-progression
oncology trial.

### 6.1 Kaplan-Meier

```{r km-table}
km_tbl |>
  flextable::flextable() |>
  flextable::theme_booktabs() |>
  flextable::autofit()
```

![KM curves with log-rank p (HR+ enjoys longer survival), 95% confidence
bands, and risk table.](../outputs/figures/tcga_km.pdf)

### 6.2 Cox proportional hazards

```{r cox-table}
cox_tbl |>
  dplyr::mutate(across(where(is.numeric), ~ signif(., 3))) |>
  flextable::flextable() |>
  flextable::theme_booktabs() |>
  flextable::autofit()
```

HR+ status reduces the hazard of death by ~42% (HR 0.58, 95% CI
0.40–0.84); each decade of age at diagnosis raises the hazard by ~28%
(HR 1.28, 95% CI 1.10–1.49). The Schoenfeld residual test flags the PH
assumption for `hr_status` (p = 0.013):

![Schoenfeld residuals.](../outputs/figures/tcga_cox_schoenfeld.pdf)

A stratified Cox model (stratifying on `hr_status` to relax the PH
assumption) was fit as a sensitivity analysis and retains a significant
age-decade effect (HR 1.29 per decade, p = 0.001).

### 6.3 Bayesian Weibull AFT — parametric cross-check

The PH violation motivates a parametric AFT model that does not require
proportional hazards. A Weibull AFT was fit in Stan (4 chains × 2,000
iterations, max R̂ = 1.003, min bulk ESS = 1,598):

```{r aft-table}
aft_tbl |>
  dplyr::filter(variable %in% c("intercept", "beta[1]", "beta[2]",
                                "time_ratio[1]", "time_ratio[2]", "shape")) |>
  dplyr::mutate(across(where(is.numeric), ~ signif(., 3))) |>
  flextable::flextable() |>
  flextable::theme_booktabs() |>
  flextable::autofit()
```

When AFT time ratios are inverted (`HR = 1 / time_ratio` is the strict
Weibull proportional-hazards correspondence, valid only under both PH
and a Weibull baseline), the Cox and Bayesian estimates agree
**directionally** but differ on the point-estimate scale: HR+ vs HR-
0.58 (Cox, 95% CI 0.40–0.84) vs 0.70 (Bayes 1/TR, 95% CrI 0.54–0.93).
The gap is the expected behavior when PH is violated — Cox estimates a
time-averaged hazard ratio while the AFT-derived HR holds only under
the parametric assumption. Both methods agree that HR+ status is
significantly protective and that each decade of age is significantly
risk-amplifying. The 95% intervals overlap modestly, not extensively,
so the agreement is informative rather than reassuring.

![Forest plot: Bayesian AFT (HR scale) vs Cox PH point estimates with
95% intervals.](../outputs/figures/tcga_aft_vs_cox_forest.pdf)

A posterior-predictive KM overlay confirms the Weibull fit visually:

![Posterior-predictive KM overlay; observed survival (dark) sits within
the envelope of 50 model-simulated replicate trials.](../outputs/figures/tcga_aft_ppc.pdf)

R̂ histogram for all model parameters:

![R̂ convergence diagnostic.](../outputs/figures/tcga_aft_rhat.pdf)

## 7. Discussion

**The adaptive design's value is operational, not statistical.** Across
non-null effects, peak power is 0.6–5.6 percentage points lower than the
fixed design. The real benefit is enrollment savings under harmful and
null scenarios: a 48% probability of stopping for futility under
HR = 1.15 and a 36% probability under the null mean the adaptive design
spares enrollment in roughly four out of ten futile trials — a
clinically and ethically relevant outcome the fixed design cannot
deliver.

**The interim is event-driven at 30% information under H1.** Triggering
the interim when 12 observed events accumulate (≈ 30% of expected events
under HR = 0.70) places the analysis inside the practical event-accrual
window for an n = 120 / 24-month trial. A 50% information target was
considered but, at the chosen sample size and event rate, almost never
accumulated before end-of-study, reducing the design to "fixed with a
near-dead futility check." Sensitivity analysis across alternative
information fractions is on the roadmap (see Limitations).

**Cox PH and Bayesian Weibull AFT directionally agree on TCGA-BRCA.**
Both methods find HR+ status significantly protective and each decade
of age significantly risk-amplifying. Point estimates differ on the HR
scale (Cox 0.58 vs Bayes 1/TR 0.70 for HR+ vs HR-) because the strict
`HR = 1/TR` correspondence holds only under both PH and Weibull
baseline assumptions; Cox estimates a time-averaged HR while the
parametric AFT does not. The agreement is therefore informative,
demonstrating that conclusions are robust to the modeling family, but
should not be over-interpreted as numerical concordance.

**Bias in the adaptive HR estimator.** Both designs produce biased
log-HR estimates, but in opposite directions:

```{r bias-table}
oc |>
  dplyr::transmute(scenario, design, hr_true,
    `bias log-HR` = sprintf("%+.3f", bias_log_hr),
    `mean HR_est` = sprintf("%.3f", mean_hr_est),
    `true HR`     = sprintf("%.2f", hr_true)
  ) |>
  dplyr::arrange(design, dplyr::desc(hr_true)) |>
  flextable::flextable() |>
  flextable::theme_booktabs() |>
  flextable::autofit() |>
  flextable::fontsize(size = 9, part = "all")
```

The fixed design shows the familiar small-sample Cox attenuation —
log-HR estimates pulled toward 0 (HR toward 1), magnitude 0.01–0.03,
direction independent of scenario. Standard, expected, and benign.

The adaptive design shows a **larger positive bias** in log-HR (0.03–0.09)
that **decreases as the true effect strengthens**. Two mechanisms
contribute:

1. **Futility-stop reporting.** When a trial stops at interim, the
   reported HR is the posterior median from the interim Bayesian fit
   (Cox PH on the few-event interim data is unstable; see R/03 comments).
   The posterior is informed by a `N(0, 1)` log-HR prior — moderately
   weak but non-trivial when only ~12 events have accumulated. The
   posterior median is therefore pulled toward HR = 1, regardless of
   the data's true direction. Under harmful HR this drags the
   distribution of reported HRs toward 1 (away from the truth of 1.15);
   under benefit, futility rarely fires, so the contribution is small.
2. **RAR allocation-imbalance under benefit.** Post-interim
   randomization allocates more to the apparently-winning arm. Under
   a true benefit, this increases events in the treatment arm
   disproportionately to control, modestly inflating the Cox HR
   estimate vs the unbiased target. The effect is bounded by the
   20/80 allocation caps but is still visible in the
   strong/very-strong scenarios.

The bias is **small relative to the effect sizes being estimated** —
roughly 5% of `log(HR_true)` for the very-strong scenario, growing
to a larger fraction under null/harmful where the truth is itself
near 1. In a real submission this magnitude is reportable but not
disqualifying; an IPTW-weighted sensitivity analysis would be the
standard companion.

**Regulatory framing.** Per FDA *Adaptive Designs for Clinical Trials of
Drugs and Biologics* (2019), §IV.A, an adaptive design submission needs
(i) pre-specified rules, (ii) Type I error control demonstrated by
simulation, (iii) bias quantified in the effect estimator. This report
provides all three: Type I is 0.019 (below the 0.025 nominal), the OBF
boundary is published in advance, and the table above is the bias
characterization required by (iii).

## 8. Limitations and design choices

- **Phase II screening design.** Maximum n = 120 with 24-month follow-up
  is deliberately small for a Phase II go/no-go trial; rpact's
  `getSampleSizeSurvival` says n ≈ 791 would be needed for 80% power at
  HR = 0.70 under this alpha-spending. Power at smaller effect sizes
  (HR 0.75 / 0.85) is correspondingly modest. This is by design, not a
  misconfiguration — a confirmatory trial would scale up.
- **Futility threshold (P(HR < 0.7 | data) < 0.20) is operator-defined.**
  A formal sensitivity analysis across alternative thresholds is on the
  roadmap.
- **Cox PH on adaptive-trial data does not adjust for RAR-induced
  allocation imbalance.** The empirical Type I (0.019) is below the
  0.025 nominal, so this is not a regulatory dealbreaker, but in a real
  submission an IPTW-weighted sensitivity analysis would accompany the
  primary Cox PH.
- **TCGA-BRCA is a toolkit validation, not a data-generating-model
  validation.** Overall survival in breast cancer differs in endpoint,
  population, and hazard shape from the simulator's hypothetical
  time-to-progression trial. The TCGA section demonstrates that the
  same Stan / KM / Cox / AFT pipeline works on real, messier data — not
  that the simulator's exponential data-generating model matches breast
  cancer biology.
- **Stan compilation in `testthat::test_dir` is fragile.** Sourcing
  rstan-heavy files repeatedly in one R process triggered "parser
  failed badly" / C-stack errors; tests therefore run each file in its
  own Rscript subprocess (`tests/testthat.R`).

## 9. SAP excerpt

A standalone mock Statistical Analysis Plan section is in
`report/sap_section.qmd` (rendered separately). It follows the standard
ICH E9-aligned outline (objectives, estimand, hypotheses, sample size,
primary analysis, missing-data handling, sensitivity analyses, safety).

## 10. Reproducibility

```bash
make sims     # runs all 12,000 trial simulations (~100 s, 4 workers)
make tcga     # fits KM, Cox, Bayes AFT on TCGA-BRCA (~30 s)
make report   # renders this document and sap_section.qmd
make all      # the lot
```

Random seeds are derived from `CONFIG$simulation$seed = 20260513`. CI runs
a reduced (`--n-sims 100`) version of the pipeline on every push.

## 11. References

- ICH E9(R1) — *Statistical Principles for Clinical Trials, Addendum on
  Estimands and Sensitivity Analyses*, 2019.
- FDA — *Adaptive Designs for Clinical Trials of Drugs and Biologics*,
  Guidance for Industry, November 2019.
- O'Brien PC, Fleming TR — A multiple testing procedure for clinical
  trials. *Biometrics* 1979; 35: 549–556.
- Cox DR — Regression models and life-tables. *J R Stat Soc B* 1972;
  34: 187–220.
- Wassmer G, Brannath W — *Group Sequential and Confirmatory Adaptive
  Designs in Clinical Trials*. Springer, 2016. (Background for `{rpact}`.)
- Stan Development Team — *Stan Reference Manual*, v2.32, 2023.
- TCGA-BRCA: The Cancer Genome Atlas Network. *Comprehensive molecular
  portraits of human breast tumours.* Nature 2012; 490: 61–70.
