---
title: "Statistical Analysis Plan (Mock Section)"
subtitle: "Phase II Adaptive Oncology Trial — Time-to-Progression Endpoint"
author: "Cris Taylor"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
  pdf:
    toc: true
    geometry: margin=1in
execute:
  echo: false
  warning: false
  message: false
---

> **Document status.** This is a mock SAP section developed for portfolio
> demonstration purposes. Cross-references such as "the protocol" or
> "Section 8.1" are illustrative rather than literal. Formal CDISC ADaM
> dataset and TLF specifications would accompany a real SAP.

## 1. Study objectives

### 1.1 Primary objective

To compare the time from randomization to disease progression or death
from any cause (progression-free survival, PFS) between patients
receiving the experimental therapy and patients receiving standard-of-care
control, in patients with the target indication.

### 1.2 Secondary objectives

- Compare overall survival between treatment arms.
- Characterize the safety profile of the experimental therapy.
- Estimate response rate and duration of response.

## 2. Primary estimand (ICH E9(R1))

The primary estimand is specified through the five attributes of the
ICH E9(R1) framework. A treatment-policy strategy is used for the most
common intercurrent event (treatment discontinuation), reflecting an
intent-to-treat philosophy appropriate for a Phase II exploratory study.

```{r estimand-sap}
suppressPackageStartupMessages({ library(flextable); library(tibble) })
tibble::tribble(
  ~Attribute,                  ~Specification,
  "Treatment",                 "Experimental therapy administered per protocol-defined schedule vs standard-of-care control, both continued until disease progression, unacceptable toxicity, or withdrawal of consent.",
  "Population",                "Patients with the target indication meeting all inclusion criteria and no exclusion criteria as specified in the protocol.",
  "Endpoint (variable)",       "Time from date of randomization to date of first documented disease progression per RECIST 1.1 (assessed by investigator) or death from any cause, whichever occurs first.",
  "Intercurrent events",       "Treatment discontinuation for any reason: treatment-policy strategy (events occurring after discontinuation are included).",
  "Population-level summary",  "Hazard ratio (experimental vs control) with 95% confidence interval, estimated from a Cox proportional-hazards model."
) |>
  flextable() |> theme_booktabs() |> autofit() |> fontsize(size = 10, part = "all")
```

## 3. Endpoints

### 3.1 Primary endpoint

**Progression-free survival (PFS)**, defined as the time from randomization
to the earlier of:

(a) first documented disease progression (RECIST 1.1, investigator-assessed); or
(b) death from any cause.

Patients without an event are censored at the last tumor assessment date
prior to data cut-off. The censoring rules follow the FDA 2018 guidance
on time-to-event endpoints in oncology.

### 3.2 Key secondary endpoints

- **Overall survival (OS)** — time from randomization to death from any cause.
- **Objective response rate (ORR)** — proportion of patients achieving
  complete or partial response per RECIST 1.1.
- **Duration of response (DoR)** — for responders only, time from first
  documented response to disease progression or death.
- **Time to treatment discontinuation (TTD)** — time from randomization
  to permanent discontinuation of experimental therapy.

### 3.3 Safety endpoints

Frequency of treatment-emergent adverse events (TEAEs) by MedDRA system
organ class and preferred term, graded by CTCAE v5.0, summarized
descriptively by treatment arm. Treatment-related TEAEs, serious AEs, and
TEAEs leading to discontinuation are summarized separately.

## 4. Statistical hypotheses

### 4.1 Primary hypothesis

Let $\lambda_E(t)$ and $\lambda_C(t)$ denote the instantaneous hazards of
PFS at time $t$ in the experimental and control arms, respectively. The
hazard ratio is $HR = \lambda_E(t) / \lambda_C(t)$, assumed constant under
proportional hazards.

- $H_0$: $HR \geq 1$ (no benefit or harm from experimental therapy)
- $H_1$: $HR < 1$ (experimental therapy reduces hazard of PFS event)

The hypothesis is tested at a **one-sided significance level of 0.025** using
a stratified log-rank test (with the Cox-PH-derived z-statistic compared
against the stage 2 O'Brien-Fleming efficacy boundary).

## 5. Analysis populations

- **Intent-to-treat (ITT)**: all randomized patients, analyzed in the
  treatment arm to which they were assigned regardless of treatment
  received. This is the primary analysis population.
- **Modified ITT (mITT)**: ITT subset of patients with at least one
  post-baseline tumor assessment. Used for sensitivity analyses.
- **Safety population**: all patients who received at least one dose of
  study treatment, analyzed by actual treatment received.

## 6. Sample-size justification

### 6.1 Phase II screening rationale

This is a **Phase II screening trial**, not a confirmatory study. The
maximum sample size of **N = 120** (60 per arm) is deliberately small
compared to the n ≈ 791 that rpact's `getSampleSizeSurvival` returns
for an 80%-powered confirmatory trial at HR = 0.70 under this
alpha-spending. The trial is sized for a go/no-go efficiency decision
under operational constraints (accrual, budget), not to definitively
reject moderate effect sizes. A confirmatory Phase III would scale up.

### 6.2 Demonstrated operating characteristics

The selected design achieves the following operating characteristics
in the simulation report (`report/index.qmd`):

- **Type I error (one-sided)**: 0.019 (adaptive) / 0.021 (fixed), both
  below the 0.025 nominal level.
- **Power across non-null effects** (adaptive design):
  - HR = 0.85: 7.5%
  - HR = 0.75: 13.9%
  - HR = 0.65: 23.8%
  - HR = 0.55: 38.7%
- **Probability of early futility stop** (adaptive design):
  - Harmful (HR = 1.15): 48%
  - Null (HR = 1.00): 36%
  - Mild benefit (HR = 0.85): 22%
  - Very strong benefit (HR = 0.55): 6%
- **Expected sample-size savings**: 5.4% under the null, 7.7% under the
  harmful direction.

The full simulation (12,000 trials across six HR scenarios) is reported
in the main simulation report.

## 7. Primary analysis method

### 7.1 Group-sequential test

The primary analysis is a Cox proportional-hazards model with treatment arm
as the sole covariate, fit at the planned final analysis (after follow-up
completion at month 24 or after all patients have been on study for at
least 12 months, whichever is later). The treatment effect is summarized
as the hazard ratio (experimental vs control) with 95% confidence interval.

The decision rule applies the O'Brien-Fleming group-sequential boundary
from a 2-stage design with one interim look at 30% information time
under the design alternative HR = 0.70, at one-sided $\alpha = 0.025$:

- **Stage 1 (interim)**: trial may stop for futility (see Section 8) but
  not for efficacy. No efficacy z-boundary is applied at interim.
- **Stage 2 (final)**: $z$-boundary = 1.969. Reject $H_0$ if the final
  Cox-derived z-statistic ($z = -\hat{\beta} / \mathrm{se}(\hat{\beta})$,
  where $\hat{\beta}$ is the log-HR estimate) exceeds the stage 2
  boundary AND the HR point estimate is below 1.

The Cox PH proportional hazards assumption is assessed via Schoenfeld
residual tests (Section 10).

## 8. Adaptive design procedures

### 8.1 Interim analysis

A single interim analysis is conducted when **12 observed events** have
accumulated — approximately 30% of the expected total events under the
design alternative (HR = 0.70). The interim therefore fires on the
basis of statistical information, not enrollment progress. The interim
analysis is performed by an independent statistician unblinded to
treatment assignment, with results reviewed by an Independent Data
Monitoring Committee (IDMC).

### 8.2 Futility stopping rule

At the interim, a Bayesian exponential survival model is fit to the
observed data:

$$
T_i \mid \text{arm}_i, \lambda_C, HR \sim
\text{Exponential}(\lambda_C \cdot HR^{1\{\text{arm}_i = E\}}),
$$

with weakly informative priors $\lambda_C \sim \text{Gamma}(2, 80)$
(centered at the assumed baseline hazard of 0.025/month) and
$\log HR \sim \mathcal{N}(0, 1)$.

The trial **stops for futility** if the interim posterior probability of
clinically meaningful benefit is below threshold:

$$
\Pr(HR < 0.70 \mid \text{interim data}) < 0.20.
$$

A futility stop results in $H_0$ being accepted; no statistical
adjustment is applied to the type I error because futility stopping is
binding.

### 8.3 Response-adaptive randomization

If the trial continues past interim, the allocation ratio is updated
from 1:1 to a Thompson-style scheme:

$$
\alpha_E = \max\left(0.20,\, \min\left(0.80,\, \sqrt{\Pr(HR < 1 \mid \text{data})}\right)\right),
$$

where $\alpha_E$ is the probability of assigning the next patient to the
experimental arm. The 20%–80% caps prevent extreme imbalance that would
inflate variance and complicate the final analysis.

Allocation is re-evaluated every 20 enrollees by refitting the Bayesian
model on accrued data. The interim posterior and each refit posterior
are documented in trial records for regulatory traceability.

### 8.4 Operating characteristics

Full operating characteristics across six HR scenarios (1.15, 1.00, 0.85,
0.75, 0.65, 0.55) are reported in `report/index.qmd` §5 and Figures
5.1–5.4. Type I error is controlled at 0.019 under the null hypothesis
(adaptive design) / 0.021 (fixed design), verified by 1,000 simulated
trials per scenario (MCSE on Type I ≈ 0.004).

## 9. Handling of missing data and intercurrent events

### 9.1 Missing data on primary endpoint

The primary endpoint (PFS) is fully observed (either as an event or as a
censoring time) by design. No imputation is performed for tumor
assessments missed within tolerance windows; censoring uses last
documented evaluable assessment.

### 9.2 Intercurrent events

Per the primary estimand (Section 2):

- **Treatment discontinuation**: treatment-policy strategy. Events
  occurring after discontinuation are included in the analysis.
- **Switching to alternative therapy**: treatment-policy strategy; data
  collection continues per protocol.
- **Death**: a component of the primary endpoint (composite endpoint
  treatment).

### 9.3 Definition of analysis cut-off

The data cut-off for the primary analysis is the date 24 months after the
first patient is randomized (or 12 months after the last patient is
randomized, whichever is later). Patients without an event at cut-off
are censored at their last tumor assessment date.

## 10. Sensitivity analyses

### 10.1 Stratified Cox PH

A stratified Cox model stratifying on baseline factors with potential PH
violation (e.g., prior systemic therapy line, performance status) is fit
as a sensitivity analysis. Results are reported alongside the primary
unstratified analysis.

### 10.2 Bayesian Weibull AFT

A parametric Weibull accelerated-failure-time model is fit to the same
data as a parametric sensitivity analysis. Posterior medians and 95%
credible intervals on the hazard ratio (computed as the inverse of the
posterior time ratio) are reported. Agreement with the Cox PH estimate
provides reassurance that the proportional-hazards assumption is not
materially violated. This dual reporting is illustrated for TCGA-BRCA
in `report/index.qmd` Section 6.3.

### 10.3 Schoenfeld diagnostic

The proportional-hazards assumption is formally tested via
`survival::cox.zph()` using scaled Schoenfeld residuals. Any covariate
with $p < 0.05$ is investigated graphically, and a stratified Cox model
is fit as the sensitivity analysis if the global test is violated.

### 10.4 Alternative estimand: hypothetical strategy

A secondary analysis under the "hypothetical strategy" (assuming
treatment discontinuation did not occur) is performed by censoring
patients at the time of discontinuation. This analysis supplements but
does not replace the primary treatment-policy analysis.

### 10.5 Sensitivity to the futility threshold

The futility rule's threshold $\theta$ (`P(HR < 0.7 \mid data) < \theta`
→ stop) is an operator-defined parameter. We assessed sensitivity by
sweeping $\theta \in \{0.10, 0.15, 0.20, 0.30\}$ across all six HR
scenarios with 300 simulated trials per cell (`R/11_futility_sensitivity.R`,
7,200 sims total, ~90 s wall time). Full results in
`outputs/tables/futility_sensitivity.csv`; key findings:

```{r futility-sens-table}
sens <- read.csv(here::here("outputs", "tables", "futility_sensitivity.csv"))
sens |>
  dplyr::transmute(
    Scenario           = scenario,
    `True HR`          = sprintf("%.2f", hr_true),
    `θ`                = sprintf("%.2f", futility_threshold),
    `Pr(reject)`       = sprintf("%.3f", reject_rate),
    `Pr(futility)`     = sprintf("%.3f", p_futility_stop),
    `E[N]`             = sprintf("%.1f", e_n)
  ) |>
  flextable::flextable() |>
  flextable::theme_booktabs() |>
  flextable::autofit() |>
  flextable::fontsize(size = 8, part = "all")
```

![Futility-stop probability and rejection rate vs the futility
threshold $\theta$, by scenario. Top panel: stopping is highly
responsive to $\theta$ under null/harmful scenarios; bottom panel:
rejection rate is essentially invariant to
$\theta$.](`r here::here('outputs','figures','futility_sensitivity.png')`)

**Key findings.**

- **Type I error is robust to the threshold.** Under the null,
  rejection rate is 0.013–0.017 across the sweep — well below the
  0.025 nominal at every threshold. Futility-only stopping is binding
  (cannot reject H~0~), so increasing $\theta$ cannot inflate Type I.
- **Power is essentially invariant to the threshold.** At very-strong
  benefit (HR = 0.55), reject rate is 39.3%–40.0% across $\theta$;
  at strong (HR = 0.65), 21.3%–22.7%. The differences are within MC
  noise for 300 sims (MC SE ~0.028).
- **Pr(futility stop) scales linearly with $\theta$** under null and
  harmful (21% → 45% under null; 31% → 57% under harmful as $\theta$
  goes 0.10 → 0.30) and modestly under benefit (very-strong: 0.7% →
  7.7%).
- **E[N] decreases with $\theta$** as more futility stops shorten
  trials. Under harmful: 114 → 109. Modest absolute savings, but
  cumulative ethical benefit across many trials in a program.

**Selected threshold ($\theta = 0.20$) is defensible.** The choice sits
near the middle of the sweep and achieves a 48% futility-stop
probability under harmful (vs 57% at $\theta = 0.30$) without any
detectable power cost vs the more conservative $\theta = 0.10$. A more
aggressive value (0.30) is plausible if maximum patient-sparing under
no benefit is prioritized over a small probability of stopping under
true (mild) benefit.

## 11. Safety analyses

Treatment-emergent AEs (TEAEs), defined as AEs with onset on or after the
first dose of study treatment, are summarized by:

- System organ class and preferred term per MedDRA (current version at
  database lock).
- Maximum severity (CTCAE v5.0).
- Treatment arm.

Safety analysis is descriptive only; no formal hypothesis tests are
performed. Key safety outputs include:

- Overall TEAE summary (any grade, grade ≥ 3, serious, fatal,
  treatment-related).
- TEAE table by SOC and PT, sorted by frequency.
- AE leading to discontinuation table.
- Laboratory shift tables (baseline to worst on-treatment grade).

A Safety Review Committee reviews aggregated safety data at the interim
analysis in parallel with the IDMC futility review.

---

## References

- **ICH E9(R1)** — *Statistical Principles for Clinical Trials, Addendum
  on Estimands and Sensitivity Analyses in Clinical Trials*, 2019.
- **FDA** — *Adaptive Designs for Clinical Trials of Drugs and Biologics*,
  Guidance for Industry, November 2019.
- **FDA** — *Clinical Trial Endpoints for the Approval of Cancer Drugs and
  Biologics*, Guidance for Industry, December 2018.
- **RECIST 1.1** — Eisenhauer EA et al. New response evaluation criteria
  in solid tumours: revised RECIST guideline (version 1.1).
  *Eur J Cancer* 2009; 45: 228–247.
- **CTCAE v5.0** — Common Terminology Criteria for Adverse Events,
  National Cancer Institute, 2017.
